---
title: "Project 2"
author: "Roger Wang"
date: "2026-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import Data

```{r cars}
library(arrow)
library(dplyr)
library(tidyr)

files <- choose.files()

basename(files) #making sure the correct files are grabbed.

set.seed(123)
#sampling 5% from each month
taxi <- bind_rows(
  lapply(seq_along(files), function(i) {
    f <- files[i]
  
    # extract month
    month_val <- as.integer(sub(".*-(\\d{2})\\.parquet", "\\1", basename(f)))
    
    
    read_parquet(f) %>%
      sample_frac(0.004, replace = FALSE) %>%
      mutate(month = month_val)
  })
)


head(taxi)
nrow(taxi) # Total of 2436130 records
summary(taxi)

```

# Adding Labels for Vendor, Rate Code, and Payment Type

```{r}
taxi <- taxi %>%
  mutate(
    # Vendor labels
    Vendor_name = case_when(
      VendorID == 1 ~ "Creative Mobile Technologies",
      VendorID == 2 ~ "Curb Mobility",
      VendorID == 6 ~ "Myle Technologies",
      VendorID == 7 ~ "Helix",
      TRUE ~ "Unknown"
    ),

    # Rate code labels
    Ratecode_name = case_when(
      RatecodeID == 1 ~ "Standard",
      RatecodeID == 2 ~ "JFK",
      RatecodeID == 3 ~ "Newark",
      RatecodeID == 4 ~ "Nassau/Westchester",
      RatecodeID == 5 ~ "Negotiated",
      RatecodeID == 6 ~ "Group ride",
      RatecodeID == 99 ~ "Unknown",
      TRUE ~ "Unknown"
    ),

    # Payment type labels
    payment_type_name = case_when(
      payment_type == 0 ~ "Flex fare",
      payment_type == 1 ~ "Credit card",
      payment_type == 2 ~ "Cash",
      payment_type == 3 ~ "No charge",
      payment_type == 4 ~ "Dispute",
      payment_type == 5 ~ "Unknown",
      payment_type == 6 ~ "Voided trip",
      TRUE ~ "Unknown"
    )
  )

head(taxi)
```

# Checking missing values
## A few variables have a large number of missing entries (46261 NAs out of 194890 records), including:passenger_count, RatecodeID, store_and_fwd_flag, congestion_surcharge, Airport_fee

```{r}
#Checking missing values
sapply(taxi, function(x) sum(is.na(x)))
```

# Section: Missing Passenger Count Analysis by Vendor

## Question: How does missing passenger count data vary across taxi vendors
## Myle Technologies has 100% missing data, which suggests a systematic issue rather than random missingness. In contrast, Creative Mobile Technologies (16.4%) and Curb Mobility (25.8%) have moderate levels of missing data, while Helix has no missing values at all.

## Question: How are passenger counts distributed across different vendors?
## Most trips across all vendors have a passenger count of 1, with higher counts (6–9) being rare outliers. Curb Mobility shows slightly more variability, but overall patterns are similar across vendors.

```{r}
taxi %>%
  count(Vendor_name, missing = is.na(passenger_count)) %>%
  pivot_wider(names_from = missing, values_from = n, values_fill = 0) %>%
  mutate(
    total = `FALSE` + `TRUE`,
    missing_pct = `TRUE` / total * 100
  )

library(ggplot2)

taxi %>%
  filter(!is.na(passenger_count)) %>%
  ggplot(aes(x = Vendor_name, y = passenger_count)) +
  geom_boxplot() +
  labs(
    title = "Passenger Count Distribution by Vendor",
    x = "Vendor",
    y = "Passenger Count"
  ) +
  theme_minimal()

taxi %>%
  filter(!is.na(passenger_count)) %>%
  ggplot(aes(x = factor(passenger_count))) +
  geom_bar() +
  facet_wrap(~Vendor_name) +
  labs(
    title = "Passenger Count Frequency by Vendor",
    x = "Passenger Count",
    y = "Count"
  ) +
  theme_minimal()
```


# Question: How should missing passenger count data be handled?
## This suggests the missingness is vendor-specific, not random. So I imputed missing passenger counts as 1 for non-Myle vendors (since 1 is most common), but avoided doing so for Myle to prevent bias.

```{r}
taxi <- taxi %>%
  mutate(
    passenger_count_missing = is.na(passenger_count),
    passenger_count = ifelse(
      is.na(passenger_count) & Vendor_name != "Myle Technologies",
      1,
      passenger_count
    )
  )

head(taxi)
```



# Section: Impact of Missing Passenger Count on Tip Analysis

## Question: Does missing passenger count data affect tip amounts?
## After removing trips with missing passenger counts, the mean and median are very close about 2.84 and 2.01.

```{r}
# Whole dataset
taxi %>%
  summarise(
    mean_tip = mean(tip_amount, na.rm = TRUE),
    median_tip = median(tip_amount, na.rm = TRUE),
    min_tip = min(tip_amount, na.rm = TRUE),
    max_tip = max(tip_amount, na.rm = TRUE)
  )

# Excluding missing passenger_count
taxi %>%
  filter(!is.na(passenger_count)) %>%
  summarise(
    mean_tip = mean(tip_amount, na.rm = TRUE),
    median_tip = median(tip_amount, na.rm = TRUE),
    min_tip = min(tip_amount, na.rm = TRUE),
    max_tip = max(tip_amount, na.rm = TRUE)
  )
```

# Section: Handling Missing Surcharge and Airport Fee Values

## Question: How prevalent are missing values in surcharge-related variables, and how should I treat them?
## Both variables have many missing values but most entries are valid. I treated missing values as no fee applied and replaced them with 0, allowing me to keep all observations without bias.
```{r}
taxi %>%
  summarise(
    congestion_non_na = sum(!is.na(congestion_surcharge)),
    congestion_na = sum(is.na(congestion_surcharge)),
    airport_non_na = sum(!is.na(Airport_fee)),
    airport_na = sum(is.na(Airport_fee))
  )

taxi <- taxi %>%
  mutate(
    congestion_surcharge = replace_na(congestion_surcharge, 0),
    Airport_fee = replace_na(Airport_fee, 0)
  )
```

# Section: Data Consistency Checks and Fare Cleaning

## Question: Are there inconsistencies between fare and total amount, and how should I clean them?
## I checked for inconsistencies by identifying records with non-positive fares or totals, which are likely invalid. I removed observations where fare_amount ≤ 0 to ensure only meaningful trips remain.
##I then calculated tip percentage using fare_amount, since total_amount includes tips and extra fees and could distort results. After cleaning, the median tip percentage is ~21% and the mean ~17.6%, though some extreme outliers remain.
##Overall, using fare_amount provides a cleaner, more consistent basis for analysis and reduces the impact of irregularities in total_amount.

```{r}
# Case 1 count
taxi %>%
  filter(fare_amount <= 0) %>%
  nrow()

# Case 2 count
taxi %>%
  filter(total_amount <= 0) %>%
  nrow()

taxi <- taxi %>% filter(fare_amount > 0)

taxi <- taxi %>% mutate(tip_percent_fare_amount = tip_amount / fare_amount)

summary(taxi)
head(taxi)
```


# Section: Tipping Behavior by Payment Type

## Question: How does average tip percentage vary by payment type, and which payment methods have the highest and lowest tipping rates?
## Answer: Tip percentages vary by payment type: credit cards have the highest (25.4%), flex fare is low (1.7%), and cash appears near zero due to unrecorded tips. “Dispute” and “No charge” are excluded as they don’t reflect real tipping behavior, leaving credit cards with clearly higher recorded tips than flex fare.

```{r}
library(ggplot2)
library(scales)

#Payment type
taxi %>%
  group_by(payment_type_name) %>%
  summarise(
    avg_tip = mean(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(payment_type_name, avg_tip), y = avg_tip, fill = payment_type_name)) +
  geom_col() +
  geom_text(aes(label = percent(avg_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Average Tip % by Payment Type",
    x = "Payment Type",
    y = "Average Tip %"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```

```{r}
#  Only keep Flex fare, Credit card
taxi <- taxi %>%
  filter(payment_type_name %in% c("Flex fare", "Credit card"))

head(taxi)

#Payment type
taxi %>%
  group_by(payment_type_name) %>%
  summarise(
    avg_tip = mean(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(payment_type_name, avg_tip), y = avg_tip, fill = payment_type_name)) +
  geom_col() +
  geom_text(aes(label = percent(avg_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Average Tip % by Payment Type",
    x = "Payment Type",
    y = "Average Tip %"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```
# Section: Tipping Behavior by Vendor

## Question: How does the median tip percentage vary across taxi vendors, and which vendor shows the highest tipping behavior?

## Anaswer: Median tip percentages differ across vendors. Helix has the highest median tip (28.2%), followed by Creative Mobile Technologies (24.1%) and Curb Mobility (23.4%). Myle Technologies shows a median tip of 0%, which likely reflects missing or unrecorded tip data rather than actual absence of tipping. Overall, Helix demonstrates the strongest tipping behavior among the vendors.

```{r}
#Vendor
taxi %>%
  group_by(Vendor_name) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(Vendor_name, median_tip), y = median_tip, fill = Vendor_name)) +
  geom_col() +
  geom_text(aes(label = percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Median Tip % by Vendor",
    x = "Vendor",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Section: Tipping and Fare Patterns by Passenger Count

## Question: How do median tip percentages and fare amounts vary with passenger count?

## Answer: Median tip percentages remain relatively stable for smaller groups (1–6 passengers), generally ranging from about 23% to 28%, with a slight peak around 5–6 passengers (~27%). However, tipping drops noticeably for larger groups (8 passengers 21.1% median tip –9 passengers 12.8% median tip).
## Fare amounts show a different pattern: most rides with 1–4 passengers have a wide distribution with many outliers, indicating variability in trip distance and cost. As passenger count increases, fare variability decreases and fares become more concentrated, though generally higher for larger groups. Overall, while fares tend to rise and stabilize with more passengers, tipping behavior weakens for larger group sizes.


```{r}
taxi %>%
  filter(
    !is.na(passenger_count),
    passenger_count > 0
  ) %>%
  group_by(passenger_count) %>%
  summarise(
    median_tip_pct = median(tip_percent_fare_amount),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = factor(passenger_count), y = median_tip_pct)) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(label = scales::percent(median_tip_pct, accuracy = 0.1)),
    vjust = -0.5,
    size = 5
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 0.35)   #increase upper limit
  ) +
  labs(
    title = "Median Tip % by Passenger Count",
    x = "Passenger Count",
    y = "Median Tip (%)"
  ) +
  theme_minimal()

taxi %>%
  filter(
    !is.na(passenger_count),
    passenger_count > 0
  ) %>%
  ggplot(aes(x = factor(passenger_count), y = fare_amount)) +
  geom_boxplot() +
  labs(
    title = "Fare Amount Distribution by Passenger Count",
    x = "Passenger Count",
    y = "Fare Amount"
  ) +
  theme_minimal()
```

```{r}
taxi %>%
  filter(
    !is.na(passenger_count),
    passenger_count > 0,
    passenger_count < 7,
    fare_amount < quantile(fare_amount, 0.95)  #remove top 1% outliers
  ) %>%
  ggplot(aes(x = factor(passenger_count), y = fare_amount)) +
  geom_boxplot() +
  labs(
    title = "Fare Distribution by Passenger Count (Zoomed)",
    x = "Passenger Count",
    y = "Fare Amount"
  ) +
  theme_minimal()


taxi %>%
  filter(
    !is.na(passenger_count),
    passenger_count > 6
  ) %>%
  ggplot(aes(x = factor(passenger_count), y = fare_amount)) +
  geom_boxplot() +
  labs(
    title = "Fare Distribution by Passenger Count (Zoomed)",
    x = "Passenger Count",
    y = "Fare Amount"
  ) +
  theme_minimal()
```

# Section: Impact of Fare Components on Tipping Behavior
## Question: How do different fare components (fare amount, extras, taxes, and surcharges) influence median tip percentages?
## Answer: Tipping decreases as fare amount increases (30.1% to 15.5%) and similarly declines with total amount (27.8% to 22.1%). In contrast, higher extras are linked to higher tips (12.4% to 27.7%). Tolls and airport fees have minimal impact. Some fees (MTA tax, improvement, congestion surcharge,  CBD congestion) lack variation, limiting insights.

```{r}
fee_columns <- c("fare_amount", "extra", "mta_tax", "tolls_amount",
          "improvement_surcharge", "total_amount",
          "congestion_surcharge", "Airport_fee", "cbd_congestion_fee")

taxi <- taxi %>%
  mutate(across(all_of(fee_columns), ~case_when(
    . <= quantile(., 0.33, na.rm = TRUE) ~ "Low",
    . <= quantile(., 0.66, na.rm = TRUE) ~ "Mid",
    . >  quantile(., 0.66, na.rm = TRUE) ~ "High"
  ), .names = "{.col}_group"))

head(taxi)

group_cols <- paste0(fee_columns, "_group")

for (col in group_cols) {
  
  plot_data <- taxi %>%
    group_by(.data[[col]]) %>%
    summarise(
      median_tip = median(tip_percent_fare_amount, na.rm = TRUE),  # ✅ changed
      .groups = "drop"
    ) %>%
    rename(group = .data[[col]]) %>%
    mutate(
      group = factor(group, levels = c("Low", "Mid", "High"))
    )
  
  p <- ggplot(plot_data, aes(x = group, y = median_tip, fill = group)) +
    geom_col() +
    geom_text(
      aes(label = scales::percent(median_tip, accuracy = 0.1)),  # ✅ updated label
      vjust = -0.3
    ) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = paste("Median Tip % (Fare) by", col),  # ✅ updated title
      x = "Group",
      y = "Median Tip %"
    ) +
    theme_minimal()
  
  print(p)
}
```


# Import borough zone file

```{r}

#Getting Borough_Zone
Borough_Zone = read.csv(file.choose(), header = TRUE, stringsAsFactors = TRUE)

head(Borough_Zone)

# Pickup
taxi <- taxi %>%
  left_join(Borough_Zone %>% select(LocationID, Borough, Zone),
            by = c("PULocationID" = "LocationID")) %>%
  mutate(pickup_location = paste(Borough, Zone, sep = " - ")) %>%
  select(-Borough, -Zone)

# Dropoff
taxi <- taxi %>%
  left_join(Borough_Zone %>% select(LocationID, Borough, Zone),
            by = c("DOLocationID" = "LocationID")) %>%
  mutate(dropoff_location = paste(Borough, Zone, sep = " - ")) %>%
  select(-Borough, -Zone)

#Checking Airport_fee > 0 when the pickup location Queens - LaGuardia Airport and Queens - JFK Airport
taxi %>% filter(!is.na(Airport_fee) & Airport_fee > 0) %>% distinct(pickup_location)

taxi <- taxi %>%
  mutate(
    airport_fee_status = case_when(
      !is.na(Airport_fee) & Airport_fee > 0 &
        !(pickup_location %in% c("Queens - LaGuardia Airport", 
                                "Queens - JFK Airport")) ~ "False Charge",
      
      !is.na(Airport_fee) & Airport_fee > 0 ~ "Valid Charge",
      
      TRUE ~ "No Charge"
    )
  )

```

# Section: Impact of Airport Fee Validity on Tipping Behavior

## Question: Does the presence of valid, invalid (false), or no airport fee affect median tip percentages?

## Answer: Median tip percentages are fairly similar across all airport fee categories, with only minor differences. Trips with a false airport charge show slightly higher median tips (26%), compared to no charge (23.8%) and valid charges (23.3%). However, the differences are small, suggesting that airport fees (valid and incorrectly applied) do not meaningfully influence tipping behavior.

```{r}
taxi %>%
  group_by(airport_fee_status) %>%
  summarise(median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
            .groups = "drop") %>%
  ggplot(aes(x = airport_fee_status, y = median_tip, fill = airport_fee_status)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  labs(
    title = "Median Tip % (Fare Amount)",
    x = "False Airport Charge",
    y = "Median Tip %"
  ) +
  theme_minimal()
```

# Section: Tipping Behavior by Pickup Borough

## Question: How does median tip percentage vary across different pickup boroughs?

## Answer: Median tip percentages vary notably by pickup borough. Manhattan (24.8%) and Queens (21.1%) show the highest tipping rates among major boroughs, indicating stronger tipping behavior in these areas. EWR (16.7%) has noticeably lower median tips. Meanwhile, Bronx, Brooklyn, and Staten Island show 0%, which likely reflects missing or unrecorded tip data rather than true absence of tipping.


```{r}
taxi %>%
  mutate(
    pickup_borough = sub(" -.*", "", pickup_location)
  ) %>%
  group_by(pickup_borough) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(pickup_borough, median_tip), y = median_tip, fill = pickup_borough)) +
  geom_col() +
  geom_text(aes(label = scales::percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Tip % by Pickup Borough",
    x = "Pickup Borough",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
# Section: Tipping Behavior by Dropoff Borough

## Question: How does median tip percentage vary across different dropoff boroughs?

## Answer: Median tip percentages vary across dropoff boroughs in a pattern similar to pickup locations. Manhattan (24.8%) and EWR (22.5%) show relatively high tipping rates, followed by Queens (11.8%) with moderate levels. Bronx and Brooklyn, Staten Island show 0%, likely due to missing or unrecorded tip data rather than actual behavior. 

```{r}
taxi %>%
  mutate(
    dropoff_borough = sub(" -.*", "", dropoff_location)
  ) %>%
  group_by(dropoff_borough) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(dropoff_borough, median_tip), y = median_tip, fill = dropoff_borough)) +
  geom_col() +
  geom_text(aes(label = scales::percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Tip % by Dropoff Borough",
    x = "Dropoff Borough",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
#checking false cbd_congestion_fee
taxi %>%
  filter(
    as.Date(tpep_pickup_datetime) < as.Date("2025-01-05") &
    !is.na(cbd_congestion_fee) &
    cbd_congestion_fee != 0
  )
```


# Section: Tipping Patterns by Pickup–Dropoff Borough Pair

## Question: How does median tip percentage vary across dropoff boroughs within each pickup borough?

## Answer: Several patterns stand out: 
### For Manhattan pickups, median tip percentages are generally highest for dropoffs to Manhattan (25.1%), EWR (23.4%), Staten Island (21.6%), and Queens (22.8%), while dropoffs to Bronx (0%) is much lower.
### For Queens pickups, tipping is also relatively strong for dropoffs to Manhattan (23.3%),  Brooklyn (19.6%), but lower for Staten Island, and EWR (8.9%) and near zero for some local categories.
### For EWR pickups, tips increase from EWR (16.7%).

## Overall, the strongest and most consistent tipping appears on trips involving Manhattan, Queens, and EWR, especially when trips connect major commercial or airport areas. In contrast, many outer-borough pickup/dropoff combinations appear sparse and may not be reliable enough for strong conclusions.


```{r}
# create borough columns once
taxi <- taxi %>%
  mutate(
    pickup_borough = sub(" -.*", "", pickup_location),
    dropoff_borough = sub(" -.*", "", dropoff_location)
  )

# loop through each pickup borough
for (p in unique(taxi$pickup_borough)) {
  
  plot_data <- taxi %>%
    filter(pickup_borough == p) %>%
    group_by(dropoff_borough) %>%
    summarise(
      median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    )
  
  p_plot <- ggplot(plot_data, aes(x = reorder(dropoff_borough, median_tip), y = median_tip)) +
    geom_col(fill = "steelblue") +
    geom_text(
      aes(label = scales::percent(median_tip, accuracy = 0.1)),
      vjust = -0.3,
      size = 3
    ) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = paste("Median Tip % | Pickup:", p),
      x = "Dropoff Borough",
      y = "Median Tip %"
    ) +
    theme_minimal()
  
  print(p_plot)
}




```

# Section: Tipping and Fare Patterns by Rate Code

## Question: How do median tip percentages and median fare amounts vary across different rate codes?

## Answer: Median tip percentages vary moderately across rate codes, while median fares differ substantially. Standard trips have the highest median tip percentage (27.0%) despite also having the lowest median fare ($12.80), suggesting riders tip a larger percentage on regular lower-cost trips. JFK and Newark trips show slightly lower median tip percentages (23.1% and 22.7%) but much higher median fares ($70.00 and $86). Negotiated (median fare $70) and Nassau/Westchester trips have the lowest nonzero median tip percentages (20.2% and 20.2%), even though Nassau/Westchester has the highest median fare ($98.90). The Unknown rate code shows a 0% median tip and a relatively low fare ($19.64), which likely reflects missing or unreliable data. Overall, higher-fare rate codes tend to have lower tip percentages, reinforcing the earlier pattern that passengers tip proportionally less on more expensive trips.


```{r}
#Rate code
taxi %>%
  group_by(Ratecode_name) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(Ratecode_name, median_tip), y = median_tip, fill = Ratecode_name)) +
  geom_col() +
  geom_text(aes(label = percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Tip % by Rate Code",
    x = "Rate Code",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")


taxi %>%
  group_by(Ratecode_name) %>%
  summarise(
    median_fare = median(fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = reorder(Ratecode_name, median_fare), y = median_fare, fill = Ratecode_name)) +
  geom_col() +
  geom_text(aes(label = scales::dollar(median_fare)), vjust = -0.3) +
  scale_y_continuous(labels = scales::dollar_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Fare Amount by Rate Code",
    x = "Rate Code",
    y = "Median Fare ($)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")

```

# Creating Time Related Features.

```{r}
library(dplyr)

taxi <- taxi %>%
  mutate(
    trip_duration_mins = as.numeric(difftime(
      tpep_dropoff_datetime,
      tpep_pickup_datetime,
      units = "mins"
    )),
    
    weekday = factor(
      weekdays(tpep_pickup_datetime),
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    ),
    
    month = factor(
      months(tpep_pickup_datetime),
      levels = month.name
    ),
    
    pickup_hour = as.integer(format(tpep_pickup_datetime, "%H")),
    
    pickup_time_of_day = case_when(
      pickup_hour >= 5  & pickup_hour < 12 ~ "Morning",
      pickup_hour >= 12 & pickup_hour < 17 ~ "Afternoon",
      pickup_hour >= 17 & pickup_hour < 21 ~ "Evening",
      TRUE ~ "Night"
    )
  )

taxi <- taxi %>%
  mutate(
    pickup_time_of_day = factor(
      pickup_time_of_day,
      levels = c("Morning", "Afternoon", "Evening", "Night")
    )
  )

```


# Section: Tipping Patterns by Pickup Borough, Dropoff Borough, and Time of Day

## Question: How does median tip percentage vary by route and time of day across pickup boroughs?

## Answer: The time-of-day breakdown generally reinforces the earlier borough-pair findings, but it also shows that tipping patterns vary meaningfully within the same route depending on when the trip occurs.

## For the major pickup boroughs:

###Tips are lower in the morning and afternoon (about 10–26%) but rise in the evening and night, often reaching 25–30%.

###From Manhattan pickups, tips are highest on longer trips, such as to Staten Island at night (~27.7%) and to unknown destinations (~25–26%), while trips to the Bronx are near 0%.

###From Queens, tips peak for unknown destinations, especially in the evening (~30.9%) and night (~26.8%), while local Queens dropoffs are much lower (around 0–7%).

###For unknown pickup locations, tips are consistently high across all times, generally around 20–28%, with peaks in the evening (~28.5%) and night (~27.4%).



```{r}
for (borough in unique(taxi$pickup_borough)) {
  plot_data <- taxi %>%
    filter(pickup_borough == borough) %>%
    group_by(dropoff_borough, pickup_time_of_day) %>%
    summarise(
      median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
      .groups = "drop"
    )

  plot_obj <- ggplot(
    plot_data,
    aes(x = dropoff_borough, y = median_tip, fill = dropoff_borough)
  ) +
    geom_col() +
    geom_text(
      aes(label = scales::percent(median_tip, accuracy = 0.1)),
      vjust = -0.3,
      size = 3
    ) +
    scale_y_continuous(
      labels = scales::percent_format(),
      expand = expansion(mult = c(0, 0.15))
    ) +
    facet_wrap(~ pickup_time_of_day) +
    labs(
      title = paste("Median Tip % | Pickup Borough:", borough),
      x = "Dropoff Borough",
      y = "Median Tip %"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1),
      legend.position = "none"
    )

  print(plot_obj)
}
```

# Section: Tipping Behavior by Weekday

## Question: How does median tip percentage vary across different days of the week?
## Answer: Median tip percentages are fairly consistent throughout the week, staying between 22.1% and 24.6%. Tips are highest on Tuesday (24.6%), followed closely by Monday and Wednesday (both 24.3%), and slightly decline toward the weekend, with the lowest on Saturday (22.1%) and Sunday (22.4%).


```{r}
taxi %>% 
  filter(trip_duration_mins > 0) %>%
  group_by(weekday) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = weekday, y = median_tip, fill = weekday)) +
  geom_col() +
  geom_text(aes(label = scales::percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Tip % by Weekday",
    x = "Weekday",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
# Section: Tipping Behavior by Month

## Question: How does median tip percentage vary across different months of the year?
## Answer: Median tip percentages are relatively stable throughout the year, mostly ranging between 23% and 25%. Tips are highest in January (25.4%) and remain strong through April (24.7%), then dip slightly in late spring and summer to around 23.1–23.4%. They rise again in the fall, reaching 24.1% in November, before dropping to the lowest point in December (21.2%).

```{r}
taxi %>% 
  group_by(month) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = month, y = median_tip, fill = factor(month))) +
  geom_col() +
  geom_text(aes(label = scales::percent(median_tip, accuracy = 0.1)), vjust = -0.3) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Tip % by Month",
    x = "Month",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


# Section: Tipping Behavior by Time of Day

## Question: How does median tip percentage vary across different times of the day?
## Answer: Median tips are highest during the afternoon (24.6%) and morning (24.5%), then drop significantly in the evening (18.4%), before slightly increasing again at night (20.0%).

```{r}
taxi %>%
  group_by(pickup_time_of_day) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = pickup_time_of_day, y = median_tip, fill = pickup_time_of_day)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Tip % by Time of Day",
    x = "Time of Day",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
# Section Title: Tipping Behavior by Payment Type

## Question : How does average tip percentage vary by payment type, and which payment methods provide meaningful insights into tipping behavior?

## Answer: Median tips are highest and most stable during the morning (about 23.4%–25.1%) and afternoon (about 22.7%–25.8%) across all weekdays.In the evening, tips drop noticeably, ranging from about 14.1% (Saturday) to 22.5% (Tuesday), with the lowest values occurring on weekends.At night, tips are slightly higher than evening but still lower than daytime, ranging from about 15.2% (Saturday) to 22.4% (Tuesday). Overall, tipping is strongest midweek and during the day, while weekends—especially Saturday evenings and nights—see the lowest tip percentages.

```{r}
taxi %>%
  group_by(weekday, pickup_time_of_day) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = weekday, y = median_tip, fill = weekday)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  facet_wrap(~ pickup_time_of_day) +
  labs(
    title = "Median Tip % by Weekday (Faceted by Time of Day)",
    x = "Weekday",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```


```{r}
taxi %>% filter(trip_duration_mins <= 0) %>% arrange(trip_duration_mins)

taxi %>% filter(trip_duration_mins > 300)%>% arrange(trip_duration_mins)
```

# Section: Tipping Behavior by Trip Duration

## Question: How does median tip percentage vary across different trip duration ranges?

## Answer: Median tip percentages are highest for short trips, peaking at 34.5% for 0–5 minutes and 29.3% for 5–10 minutes. Tips then decline as trip duration increases, dropping to 23.5% (10–20 min) and 17.9% (20–30 min). Although there is a slight increase for 30–60 minutes (18.9%), the longest trips (60+ minutes) have the lowest tips at 14.3%, showing that shorter trips tend to receive higher tip percentages.


```{r}
taxi %>%
  mutate(duration_bucket = case_when(
    trip_duration_mins <= 0 ~ "≤0",
    trip_duration_mins <= 5 ~ "0–5",
    trip_duration_mins <= 10 ~ "5–10",
    trip_duration_mins <= 20 ~ "10–20",
    trip_duration_mins <= 30 ~ "20–30",
    trip_duration_mins <= 60 ~ "30–60",
    TRUE ~ "60+"
  )) %>%
  group_by(duration_bucket) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(duration_bucket = factor(duration_bucket,
    levels = c("≤0","0–5","5–10","10–20","20–30","30–60","60+")
  )) %>%
  ggplot(aes(x = duration_bucket, y = median_tip, fill = duration_bucket)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Tip % by Trip Duration",
    x = "Trip Duration (minutes)",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
# Section: Fare Amount by Trip Duration

## Question: How does median fare amount vary across different trip duration ranges?

## Answer: Median fares increase steadily with trip duration, rising from $5.80 (0–5 min) and $9.30 (5–10 min) to $14.90 (10–20 min) and $24.70 (20–30 min). Longer trips see much higher fares, reaching $43.50 (30–60 min) and peaking at $70.00 (60+ min), showing a strong positive relationship between trip length and fare.

```{r}
taxi %>%
  mutate(duration_bucket = case_when(
    trip_duration_mins <= 0 ~ "≤0",
    trip_duration_mins <= 5 ~ "0–5",
    trip_duration_mins <= 10 ~ "5–10",
    trip_duration_mins <= 20 ~ "10–20",
    trip_duration_mins <= 30 ~ "20–30",
    trip_duration_mins <= 60 ~ "30–60",
    TRUE ~ "60+"
  )) %>%
  group_by(duration_bucket) %>%
  summarise(
    median_fare = median(fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(duration_bucket = factor(duration_bucket,
    levels = c("≤0","0–5","5–10","10–20","20–30","30–60","60+")
  )) %>%
  ggplot(aes(x = duration_bucket, y = median_fare, fill = duration_bucket)) +
  geom_col() +
  geom_text(
    aes(label = scales::dollar(median_fare)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::dollar_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Fare by Trip Duration",
    x = "Trip Duration (minutes)",
    y = "Median Fare ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Section: Tipping Behavior by Trip Distance

## Question: How does median tip percentage vary across different trip distance ranges?

## Answer: Median tip percentages are highest for very short trips, peaking at 31.1% for 0–1 miles and 26.7% for 1–2 miles. Tips then decline with distance, dropping to 19.8% (2–5 miles) and reaching the lowest at 11.3% (5–10 miles). For longer trips, tips recover slightly to around 20.8% (10–20 miles) and 20.5% (20+ miles), but remain below the highest short-distance levels.


```{r}
taxi %>%
  mutate(distance_bucket = case_when(
    trip_distance <= 0 ~ "≤0",
    trip_distance <= 1 ~ "0–1",
    trip_distance <= 2 ~ "1–2",
    trip_distance <= 5 ~ "2–5",
    trip_distance <= 10 ~ "5–10",
    trip_distance <= 20 ~ "10–20",
    TRUE ~ "20+"
  )) %>%
  group_by(distance_bucket) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(distance_bucket = factor(distance_bucket,
    levels = c("≤0","0–1","1–2","2–5","5–10","10–20","20+")
  )) %>%
  ggplot(aes(x = distance_bucket, y = median_tip, fill = distance_bucket)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Tip % by Trip Distance",
    x = "Trip Distance",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

# Section: Fare Amount by Trip Distance

## Question: How does median fare amount vary across different trip distance ranges?

## Answer: Median fares increase as trip distance grows, starting at $7.20 (0–1 miles) and $11.40 (1–2 miles), then rising to $19.10 (2–5 miles) and $33.10 (5–10 miles). Longer trips have much higher fares, reaching $60.40 (10–20 miles) and peaking at $70.00 (20+ miles), showing a strong positive relationship between distance and fare.

```{r}
taxi %>%
  mutate(distance_bucket = case_when(
    trip_distance <= 0 ~ "≤0",
    trip_distance <= 1 ~ "0–1",
    trip_distance <= 2 ~ "1–2",
    trip_distance <= 5 ~ "2–5",
    trip_distance <= 10 ~ "5–10",
    trip_distance <= 20 ~ "10–20",
    TRUE ~ "20+"
  )) %>%
  group_by(distance_bucket) %>%
  summarise(
    median_fare = median(fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(distance_bucket = factor(distance_bucket,
    levels = c("≤0","0–1","1–2","2–5","5–10","10–20","20+")
  )) %>%
  ggplot(aes(x = distance_bucket, y = median_fare, fill = distance_bucket)) +
  geom_col() +
  geom_text(
    aes(label = scales::dollar(median_fare)),
    vjust = -0.3
  ) +
  scale_y_continuous(
    labels = scales::dollar_format(),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Median Fare by Trip Distance",
    x = "Trip Distance",
    y = "Median Fare ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
taxi %>%
  filter(trip_distance > 0) %>%
  mutate(distance_bucket = case_when(
    trip_distance <= 1 ~ "0–1",
    trip_distance <= 2 ~ "1–2",
    trip_distance <= 5 ~ "2–5",
    trip_distance <= 10 ~ "5–10",
    trip_distance <= 20 ~ "10–20",
    TRUE ~ "20+"
  )) %>%
  group_by(payment_type_name, distance_bucket) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(distance_bucket = factor(distance_bucket,
    levels = c("0–1","1–2","2–5","5–10","10–20","20+")
  )) %>%
  ggplot(aes(x = distance_bucket, y = median_tip, fill = payment_type_name)) +
  geom_col(position = "dodge") +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Tip % by Distance and Payment Type",
    x = "Trip Distance",
    y = "Median Tip %"
  ) +
  theme_minimal()
```

# Section: Tipping Behavior by Trip Distance and Passenger Count

## Question: How does median tip percentage vary by trip distance when broken down by passenger count?

## Answer: 
### Across all passenger counts, tip percentages are highest for short trips and decline as distance increases. For example, with 1 passenger, tips drop from 31.1% (0–1 miles) to 3.7% (5–10 miles) before slightly rising again for longer trips (~18–20%).
### This pattern is consistent across groups: for 2–6 passengers, short trips (0–1 miles) have the highest tips (~32–34%), while mid-distance trips (5–10 miles) are lower (~23%), and longer trips (10+ miles) stabilize around 21–23%.
### Overall, regardless of passenger count, shorter trips receive higher tip percentages, while longer trips tend to have lower but more consistent tipping.

```{r}
taxi %>%
  filter(trip_distance > 0, passenger_count > 0) %>%
  mutate(distance_bucket = case_when(
    trip_distance <= 1 ~ "0–1",
    trip_distance <= 2 ~ "1–2",
    trip_distance <= 5 ~ "2–5",
    trip_distance <= 10 ~ "5–10",
    trip_distance <= 20 ~ "10–20",
    TRUE ~ "20+"
  )) %>%
  group_by(passenger_count, distance_bucket) %>%
  summarise(
    median_tip = median(tip_percent_fare_amount, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(distance_bucket = factor(distance_bucket,
    levels = c("0–1","1–2","2–5","5–10","10–20","20+")
  )) %>%
  ggplot(aes(x = distance_bucket, y = median_tip, fill = distance_bucket)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(median_tip, accuracy = 0.1)),
    vjust = -0.3,
    size = 3
  ) +
  scale_y_continuous(labels = scales::percent_format(),
                     expand = expansion(mult = c(0, 0.3))) +
  facet_wrap(~ passenger_count) +
  labs(
    title = "Median Tip % by Distance (Faceted by Passenger Count)",
    x = "Trip Distance",
    y = "Median Tip %"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


# Section: Relationship Between Tip Percentage and Trip Duration

## Question: What is the relationship between tip percentage and trip duration?
## Answer: Both plots show a clear negative relationship between trip duration and tip percentage. Short trips have higher and more variable tips, while longer trips cluster at lower, more consistent tip levels.

```{r}
taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    #fare_amount > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 4
  ) %>%
  filter(
    trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE)
    #fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = trip_duration_mins, y = tip_percent_fare_amount)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth( method = 'lm')+
  labs(
    title = "Tip % vs Trip Duration",
    x = "Trip Duration (minutes)",
    y = "Tip %"
  ) +
  theme_minimal()


taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    #fare_amount > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 4
  ) %>%
  filter(
    trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE)
    #fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = log1p(trip_duration_mins), y = tip_percent_fare_amount)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth( method = 'lm')+
  labs(
    title = "Tip % vs Log-Transformed Trip Duration",
    x = "Log(1 + Trip Duration (minutes))",
    y = "Tip %"
  ) +
  theme_minimal()
```
# Section: Relationship Between Tip Percentage and Trip Distance

## Question: What is the relationship between tip percentage and trip distance?
## Answer: Both plots show a negative relationship between trip distance and tip percentage. Short-distance trips have higher and more variable tips, while longer trips have lower, more consistent tip percentages, with the trend steadily decreasing as distance increases.

```{r}
taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    #fare_amount > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 4
  ) %>%
  filter(
    trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE)
    #fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = trip_distance, y = tip_percent_fare_amount)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth( method = 'lm')+
  labs(
    title = "Tip % vs Trip Distance",
    x = "Trip Distance (miles)",
    y = "Tip %"
  ) +
  theme_minimal()


taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    #fare_amount > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 4
  ) %>%
  filter(
    trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE)
    #fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = log1p(trip_distance), y = tip_percent_fare_amount)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth( method = 'lm')+
  labs(
    title = "Tip % vs Log-Transformed Trip Distance",
    x = "Log(1 + Trip Distance(miles))",
    y = "Tip %"
  ) +
  theme_minimal()
```
# Section: Relationship Between Tip Percentage and Fare Amount

## Question: What is the relationship between tip percentage and fare amount?
## Answer: There is a negative relationship between (log) fare amount and tip percentage. As fares increase (even on a log scale), tip percentages gradually decrease, with higher variability at lower fares and more consistent, lower tips at higher fares.

```{r}
taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 4
  ) %>%
  filter(
    #trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE)
    #fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = log1p(fare_amount), y = tip_percent_fare_amount)) +
  geom_jitter(alpha = 0.2) +
  geom_smooth( method = 'lm')+
  labs(
    title = "Tip % vs Fare Amount",
    x = "Fare Amount",
    y = "Tip %"
  ) +
  theme_minimal()
```
# Section: Predictive Performance of the Tip Percentage Model

## Question: How well does the linear regression model predict tip percentage withour removing outliers?
## Answer: The linear regression model shows moderate predictive accuracy. On the test set, it achieves an RMSE of about 0.129 and an MAE of about 0.068. This means the model’s predictions deviate from the actual tip percentage by about 6.9 percentage points on average, with larger errors reflected in the RMSE. Overall, the model captures some of the variation in tipping behavior, but prediction error remains substantial, suggesting that tip percentage is influenced by additional nonlinear or unobserved factors not fully captured by the model.

```{r}
set.seed(123)

train_idx <- sample(seq_len(nrow(taxi)), size = 0.8 * nrow(taxi))

summary(taxi)

train <- taxi[train_idx, ]
test  <- taxi[-train_idx, ]

model <- lm(
  tip_percent_fare_amount ~ 
    trip_distance +
    trip_duration_mins +
    fare_amount +
    payment_type_name +
    pickup_time_of_day +
    weekday +
    pickup_borough +
    dropoff_borough +
    Ratecode_name,
  data = train
)

#summary(model)
#confint(model)

pred <- predict(model, newdata = test)

library(caret)

RMSE(pred, test$tip_percent_fare_amount)
MAE(pred, test$tip_percent_fare_amount)
```

## Section: Model Performance After Data Cleaning

## Question: How does the linear regression model perform after cleaning the data?
## Answer: After removing outliers and restricting tip percentages to a reasonable range, the model’s performance improves substantially. The test set results show an RMSE of about 0.0977 and an MAE of about 0.0669, indicating much smaller prediction errors compared to the earlier model.


```{r}
taxi_clean <- taxi %>%
  filter(
    trip_distance > 0,
    trip_duration_mins > 0,
    fare_amount > 0,
    tip_percent_fare_amount >= 0,
    tip_percent_fare_amount <= 1
  ) %>%
  filter(
    trip_distance <= quantile(trip_distance, 0.99, na.rm = TRUE),
    trip_duration_mins <= quantile(trip_duration_mins, 0.99, na.rm = TRUE),
    fare_amount <= quantile(fare_amount, 0.99, na.rm = TRUE)
  )

summary(taxi_clean)

set.seed(123)

train_idx_clean <- sample(seq_len(nrow(taxi_clean)), size = 0.8 * nrow(taxi_clean))

train_clean <- taxi_clean[train_idx_clean, ]
test_clean  <- taxi_clean[-train_idx_clean, ]


#model_clean <- lm(
#  tip_percent_fare_amount ~ 
#    trip_distance +
#    trip_duration_mins +
#    fare_amount +
#    payment_type_name +
#    pickup_time_of_day +
#    weekday +
#    pickup_borough +
#    dropoff_borough +
#    Ratecode_name,
#  data = train_clean
#)

#summary(model_clean)
#confint(model_clean)

#pred_clean <- predict(model_clean, newdata = test_clean)

#RMSE(pred_clean, test_clean$tip_percent_fare_amount)
#MAE(pred_clean, test_clean$tip_percent_fare_amount)

# Train predictions
#pred_train <- predict(model_clean, newdata = train_clean)

# Train metrics
#RMSE(pred_train, train_clean$tip_percent_fare_amount)
#MAE(pred_train, train_clean$tip_percent_fare_amount)

```




```{r}
model_improved <- lm(
  tip_percent_fare_amount ~
    #log1p(trip_distance) +
    log1p(trip_duration_mins) + I(log1p(trip_duration_mins)^2) +
    #pickup_time_of_day + weekday +
    #pickup_borough +
    #dropoff_borough, 
    Ratecode_name,
  data = train_clean
)
    #pickup_time_of_day + weekday +
    #pickup_borough +
    #dropoff_borough, 
    #Ratecode_name,
#log1p(fare_amount) +
    #payment_type_name +

pred_clean_improved <- predict(model_improved, newdata = test_clean)

RMSE(pred_clean_improved, test_clean$tip_percent_fare_amount)
MAE(pred_clean_improved, test_clean$tip_percent_fare_amount)


# Train predictions
pred_train_improved <- predict(model_improved, newdata = train_clean)

# Train metrics
RMSE(pred_train_improved, train_clean$tip_percent_fare_amount)
MAE(pred_train_improved, train_clean$tip_percent_fare_amount)

```

